\documentclass[11pt,a4paper]{article}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage[a4paper,margin=2.4cm]{geometry}
\usepackage{enumitem}
\usepackage[autostyle]{csquotes}
\usepackage[hidelinks]{hyperref}
\usepackage[backend=biber,style=authoryear,maxcitenames=2,maxbibnames=12]{biblatex}
\addbibresource{../bibliographie/references.bib}

\setlength{\parindent}{0pt}
\setlength{\parskip}{0.65em}
\setlist[itemize]{leftmargin=1.5em,itemsep=0.35em,topsep=0.35em}
\setlist[enumerate]{leftmargin=1.8em,itemsep=0.5em,topsep=0.4em}

\newcommand{\projectversion}{2}
\newcommand{\projectdate}{12 July 2026}

\title{The Emergence of the Interesting\\
\large Birth of Ideas, Understanding, and Singularity of Forms}
\author{François Pachet}
\date{Thesis project -- version \projectversion\\\small \projectdate}

\begin{document}

\maketitle

\begin{abstract}
This project aims to establish the interesting as a philosophical object. It
starts from an ordinary yet elusive experience: an idea emerges, a form becomes
intelligible, a passage holds our attention and calls for repetition. Its central
claim is that these events are neither isolated properties of objects nor merely
private preferences. They arise from a temporal relation between an organised
form and a subject with a perceptual history. Music is the main laboratory of
this inquiry; forty years spent building artificial-intelligence systems provide
a second, both experimental and reflexive, field. The aim is to articulate
phenomenology, philosophy of form, aesthetics, cognitive science, and
computational modelling without reducing the experience of the interesting to
any one of these registers.
\end{abstract}

\section{Origin and Aim of the Project}

This work emerges from a constructivist and empirical practice of artificial
intelligence research. I have spent forty years building systems that generate,
transform, or accompany music and text. The doctoral project is not intended to
summarise this work or retrospectively impose an artificial philosophical unity
on it. It starts from a more demanding question: \emph{what am I doing} when I
build a system, recognise a solution, reject an output that is correct but
uninteresting, or stop before a form that seems to impose itself?

This inquiry intersects problems that philosophy has treated under different
names: access to the intelligible in Plato and Aristotle
\parencite{platon2016republique,platon1995theetete,aristote1991metaphysique,aristote2005seconds};
intuition and evidence in Descartes and Spinoza
\parencite{descartes2002regles,descartes1992meditations,spinoza1966ethique};
judgement in Kant \parencite{kant2001raison,kant1993juger}; invention and the
reconfiguration of problems in Peirce, Poincaré, Hadamard, Bachelard, and Valéry
\parencite{peirce1955writings,poincare1908science,hadamard1959invention,bachelard1999formation,valery1973cahiers};
memory, duration, and the temporal constitution of meaning in Bergson, Husserl,
and Merleau-Ponty
\parencite{bergson2013evolution,bergson2013pensee,husserl1970experience,husserl1957logique,merleau1945phenomenologie};
aspect change in Wittgenstein \parencite{wittgenstein2004recherches}; and
symbolisation, difference, and repetition in Goodman and Deleuze
\parencite{goodman1976languages,deleuze1968difference}.

The unifying hypothesis is that the birth of an idea, the experience of
understanding, and fascination with a form are three moments of the same
phenomenon: the emergence of something that stands out for a subject as an
organisation that is both intelligible and singular.

\section{Guiding Question and Central Claim}

The guiding question is: \emph{under what conditions does something appear
interesting to a subject?} A song, a proof, a sentence, or an idea may hold our
attention disproportionately, provoke repetitions and variations, then lose or
transform its power. This effect cannot be reduced to raw novelty, because noise
surprises without necessarily being interesting; nor to beauty, because a
familiar form may remain beautiful while becoming predictable; nor to
popularity; nor to subjective preference understood as a datum without a
history.

The central claim is:

\begin{quote}
The interesting is a relational and temporal property. It emerges when an
organised form transforms the state of attention, memory, or understanding of a
historically, affectively, bodily, and culturally constituted subject.
\end{quote}

Relativity to a subject does not entail that all judgements are equivalent. A
relation can be described: we can identify the expectations made possible by an
object, the competencies required to experience them, the transformations that
renew attention, and the variants that would destroy the effect. The structure
of the object and the history of the subject are therefore two irreducible
conditions of the same encounter.

\section{Working Hypotheses}

\begin{enumerate}
  \item \textbf{The interesting is a process, not a label.} For a temporal
  object, the relevant unit is a trajectory made of preparation, expectation,
  deviation, resolution, reinterpretation, and sometimes boredom. The same
  event changes value according to its position and to what later events make of
  it.

  \item \textbf{An interesting form may invent the problem it solves.} Some
  solutions appear necessary after they emerge even though they could not have
  been deduced beforehand. They reveal their own constraints and create an
  impression of surface simplicity that conceals an improbable genesis. The
  relevant rarity then lies less in raw statistical frequency than in the small
  class of solutions to an implicit problem.

  \item \textbf{Interest moves with learning.} A pattern that is initially
  opaque may become accessible, produce a gain in understanding, and then lose
  its force once it has been acquired. Compression-progress theory formalises
  one version of this dynamic
  \parencite{schmidhuber1997interesting,schmidhuber2009compression}. It must,
  however, be understood relative to an observer's computational capacities,
  memory, and culture, rather than as a universal measure of an object.

  \item \textbf{The interesting often lies on a moving frontier of the almost
  learnable.} What has already been learned is trivial; noise remains
  incompressible; a structure too remote from the subject's capacities remains
  opaque. A form is interesting when it promises an accessible transformation
  at the cost of non-zero effort. This claim extends, while complicating, the
  inverted-U curve of experimental aesthetics
  \parencite{berlyne1974studies,abdallah2009information}.

  \item \textbf{Attention is active and fallible.} Listeners internally vary,
  reconstruct, and compare what they hear. This activity may disclose a genuine
  organisation, but it may also manufacture the interest it claims to discover.
  Familiarity, the prestige of an author, and the desire to solve an enigma are
  possible sources of false positives. A theory of the interesting must
  therefore contain its own critique.
\end{enumerate}

\section{Theoretical Framework: Subject, Form, and Emergence}

The theoretical framework will articulate three requirements. The first comes
from philosophy of the subject: the experience of the interesting depends on
conditions of reception and judgement without being reducible to an objective
feature. The second comes from a phenomenology of temporality: retention,
protention, bodily memory, and aspect change describe how a form is constituted
through time. The third comes from morphodynamic theories and complex systems:
a global organisation has properties that cannot be read from its isolated
elements.

This framework takes up the attempt to naturalise certain transcendental
problems without eliminating their phenomenal dimension
\parencite{petitot1993phenomenologie,petitot2002naturaliser}. It does not assume
that scientific modelling will exhaust the interesting. It seeks a controlled
relation between first-person descriptions, formal analyses, and testable
mechanisms. The inquiry will distinguish possible invariants of attention from
habits limited to a context, notably those of Western tonal music.

\section{Music as a Philosophical Laboratory}

Music makes the relations between memory, expectation, surprise, and resolution
especially visible. Meyer and later Narmour showed that musical affect and
meaning depend on learned expectations that are fulfilled, delayed, or thwarted
\parencite{meyer1956emotion,narmour1990basic,narmour1992complexity}.
Information-dynamic and neuro-symbolic models offer ways of specifying these
processes \parencite{abdallah2009information,berger1999expectations,gang1999unified}.

Yet explaining why a track is interesting cannot amount to locating an average
quantity of surprise. The inquiry will address trajectories of attention at
several scales: local micro-emotions, relations between melody and harmony,
global form, listening history, and the listener's own activity. The Jotney
regime provides an initial case: melodic autonomy and harmonic mobility create a
narrow band in which surprise becomes retrospectively necessary.

\emph{Histoire d'une oreille} provides a phenomenological vocabulary and a
method \parencite{pachet2018oreille}. The \emph{escatte}, \emph{escassepte},
\emph{misette}, \emph{neuvéclamine}, \emph{cébimine}, and \emph{stabilème} are
not new names for chords; they describe attentional transformations produced by
temporal relations. \emph{Superimposed listening} submits music to internal
variations. An \emph{optale} is the feeling that a form is a local optimum
because nearby variants fail; a \emph{pseudoptale} warns that familiarity may
simulate this necessity. \emph{False evidence} distinguishes a banal phrase
that is retrospectively requalified from a merely \emph{necessary substrate}.
Finally, \emph{overattentive listening} names the risk that listeners themselves
produce the resolution they attribute to the work.

These concepts make it possible to move from ``I like this track'' to comparable
claims: what transformation occurs, at what moment, through what preparation,
for which ear, and what would become of it under a controlled variation? They are
not intended to form a definitive dictionary. They are instruments for
describing and testing a theory of musical attention.

\section{Scientific Practice and Philosophical Inquiry}

AI systems constitute a second laboratory. Work on Flow Machines, reflexive
interactions, Markov generation under constraints, and computational
interestingness has produced objects, protocols, successes, and failures
\parencite{pachet2006clefs,pachet2008machines,pachet2012virtuosity,pachet2021assisted,pachet2026markov}.
It shows, in particular, that local forward generation does not always suffice
to satisfy a global organisation: producing an interesting form often requires
sampling to be articulated with constraint solving.

Three levels must remain distinct:

\begin{itemize}
  \item scientific results and effective properties of systems;
  \item philosophical claims about invention, attention, or value;
  \item argued articulations explaining what the construction of a system makes
        thinkable without claiming that the system proves it.
\end{itemize}

This articulation is itself an expected contribution of the thesis. A
computational model may make a hypothesis explicit, produce a counterexample, or
reveal that a notion is too general. It does not suffice to naturalise the
experience it schematises.

\section{Method and Corpus}

The research will combine:

\begin{itemize}
  \item a conceptual analysis of the interesting and its neighbours: the new,
        the surprising, the beautiful, the relevant, the difficult, the rare,
        and the understandable;
  \item a study of philosophical traditions concerned with form, judgement,
        temporality, invention, and emergence;
  \item a situated phenomenology of musical, literary, mathematical, and
        scientific cases, including the history of the formation of an ear;
  \item counterfactual comparisons: internal variations, merely competent
        continuations, failed covers, or failed imitations;
  \item a reflexive analysis of constructed systems and, where relevant, new
        experiments on attention, repetition, abandonment, and the evolution of
        judgements;
  \item a confrontation with contemporary models of expectation, compression,
        curiosity, and constrained generation.
\end{itemize}

The main corpus comprises the two essays on the impossibility of creation and on
virtuosity \parencite{pachet2026impossibilite,pachet2026virtuosite},
\emph{Histoire d'une oreille}, the work on Jotney songs, scientific projects on
interestingness and generative systems, and the philosophical and scientific
sources listed in the bibliography. Older cases will be retained only when they
directly illuminate the question of the interesting.

\section{Expected Contributions}

The thesis aims at five contributions:

\begin{enumerate}
  \item a relational and temporal definition of the interesting, distinct from
        novelty, beauty, and preference;
  \item a theory of invented necessity linking form, implicit problem, rarity of
        solutions, and improbable genesis;
  \item a theory of musical attention articulating micro-emotions,
        expectations, global form, learning, and listener activity;
  \item a method for moving between scientific construction and philosophical
        claims, with explicit criteria of scope and limitation;
  \item a set of falsifiable, or at least comparable, hypotheses concerning
        trajectories of interest, their transformations, and their false
        positives.
\end{enumerate}

\section{Provisional Outline}

\begin{enumerate}
  \item \textbf{Distinguishing the interesting}: concept, neighbouring notions,
        history of the problem, and objections.
  \item \textbf{Emergence and understanding}: appearance of a form, aspect
        change, and birth of problems.
  \item \textbf{Form, memory, and attention}: temporality, learning, the almost
        learnable, and exhaustion of an effect.
  \item \textbf{Invented necessity}: constraints, rarity, self-stability,
        difficulty of production, and surface simplicity.
  \item \textbf{Why is this track interesting?}: musical expectations,
        micro-emotions, optales, global form, and overattention.
  \item \textbf{Building in order to understand}: AI systems, experiments,
        philosophical scope, and limits of naturalisation.
\end{enumerate}

\section{Directly Related Personal Work}

\begin{itemize}
  \item \textcite{pachet2026virtuosite} analyses forms of virtuosity and their
        intended effects on listeners.
  \item \textcite{pachet2026impossibilite} examines scientific and conceptual
        obstacles to the idea of creating something new.
  \item \textcite{pachet2026markov} synthesises work on Markov modelling and
        constraint-controlled generation.
  \item \textcite{pachet2021assisted} derives new categories of the new from the
        Flow Machines projects.
  \item \textcite{pachet2018oreille} describes the ontogenesis of a musical ear
        and provides the attentional concepts used here.
  \item \textcite{pachet2012virtuosity} connects virtuosity, creativity, and the
        construction of the Virtuoso system.
  \item \textcite{pachet2008machines} presents reflexive systems as machines for
        stepping outside oneself.
  \item \textcite{pachet2006clefs} offers an initial model of interesting
        melodies.
\end{itemize}

\printbibliography[title={References}]

\end{document}
